---
title: "team_plots"
author: "Rebeka Róth"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(ggplot2)
library(dplyr)
library(stringr)
library(tidyr)
library(fmsb)
library(scales)
library(ggforce)
library(ggshakeR)
```

```{r}
teams = c("Brown Bears", "Columbia Lions", "Cornell Big Red", "Dartmouth Big Green", "Harvard Crimson", "Penn Quakers", "Princeton Tigers", "Yale Bulldogs")

seasons = c("2021", "2022", "2023", "2024")
```

```{r}
dt = data.frame()
col_names = c("date", "match", "comp", "duration", "team", "scheme", "goals", "xg", "shots", "shots_on_target", "shots_pct", "passes", "passes_accurate", "pass_pct", "poss_pct", "losses", "losses_low", "losses_med", "losses_high", "rec", "rec_low", "rec_med", "rec_high", "duels", "duels_won", "duels_pct")

# example file location and name: /Users/rothrebeka/Documents/OneDrive - Yale University/Spring 2025/senior project/data/team stats/Team Stats Brown Bears 2021.xlsx

for (team in teams) {
  for (season in seasons) {
    file = paste0("/Users/rothrebeka/Documents/OneDrive - Yale University/Spring 2025/senior project/data/team stats/Team Stats ", team, " ", season, ".xlsx")
    data = read_excel(file)
    
    # get rid of first two rows
    data = data[-c(1, 2), ]
    
    # apply column names
    colnames(data) = col_names
    data$season = season
    dt = rbind(dt, data)
  }
}

# Remove duplicates
dt = dt %>% unique()

```


```{r}
# Create new columns (opp.goals, opp.xg, etc)
# Make them be empty
# Take odd and even rows and fill them with appropaite values such as df[odd.rows, opp.cols] = df[even.rows, team.cols]

team_cols = c("goals", "xg", "shots", "shots_on_target", "shots_pct", "passes", "passes_accurate", "pass_pct", "poss_pct", "losses", "losses_low", "losses_med", "losses_high", "rec", "rec_low", "rec_med", "rec_high", "duels", "duels_won", "duels_pct")
opp_cols = paste0("opp.", team_cols)

dt[opp_cols] = NA

odd_rows = seq(1, nrow(dt), by = 2)
even_rows = seq(2, nrow(dt), by = 2)

dt[odd_rows, opp_cols] = dt[even_rows, team_cols]
dt[even_rows, opp_cols] = dt[odd_rows, team_cols]

```

```{r}
# Create new columns for home team and away team
# split on hyphen in match column
# get rid of result (everything after last space)
# team before hyphen is home team, team after hyphen is away team

dt = dt %>%
  mutate(match = str_remove(match, "\\s\\d+:\\d+$")) %>% # Remove result
  separate(match, into = c("home_team", "away_team"), sep = " - ")
```

Filter for teams that are in Ivy
```{r}
dt = dt %>%
  filter(team %in% teams)
```

*Bar Charts*

Create bar charts for the following variables:
goals
xG
shots on target
pass percentage 
possession percentage
losses low third
recoveries high third
duels won percentage
opponent goals
opponent xg 
opponent shots on target 

All bar charts should include averages for all Ivy 8 teams. Make sure to color Yale's chart Blue, other teams charts should be grey. Add label to each bar as well as the actual value.

```{r}
# Create a function that creates a bar chart for a given variable
# Function to create bar charts with averaging options

# Function to create horizontal bar charts with Yale in blue
plot_bar_chart <- function(df, var, var_label, seasons = "2024", n_games=NULL) {

  if (is.null(n_games)){
    df <- df %>%
      filter(season == seasons)
  } else {
    df <- df %>%
      group_by(team) %>%
      top_n(n_games, wt = date)
  }
  
  avg <- df %>%
  group_by(team) %>%
  summarise(avg = mean(!!sym(var), na.rm = TRUE))
  
  
  # Plot
  ggplot(avg, aes(x = avg, y = reorder(team, avg), fill = team)) +
    geom_bar(stat = "identity") +
    geom_text(aes(label = round(avg, 2)), hjust = -0.1, size = 3) +
    labs(title = paste0("Average ", var_label, " by Team"),
         x = var_label,
         y = "Team") +
    theme_minimal() +
    theme(legend.position = "none") +
    scale_fill_manual(values = c("Brown Bears" = "grey", "Columbia Lions" = "grey", "Cornell Big Red" = "grey", "Dartmouth Big Green" = "grey", "Harvard Crimson" = "grey", "Penn Quakers" = "grey", "Princeton Tigers" = "grey", "Yale Bulldogs" = "#0f4d92"))
}

```

```{r}
# Example Call
plot_bar_chart(dt, "goals", "Goals", season="2022", n_games=10)
```

```{r}
vars <- c("goals", "xg", "shots_on_target", "pass_pct", "poss_pct",
          "losses_low", "rec_high", "duels_pct",
          "opp.goals", "opp.xg", "opp.shots_on_target")

var_labels <- c("Goals", "Expected Goals (xG)", "Shots on Target", "Pass Percentage", 
                "Possession Percentage", "Losses in Low Third", "Recoveries in High Third", 
                "Duels Won Percentage", "Opponent Goals", "Opponent xG", "Opponent Shots on Target")

# create bar charts for all variables
for (i in 1:length(vars)) {
  plot = plot_bar_chart(dt, vars[i], var_labels[i], seasons="2021", n_games=10)
  print(plot)
}
```


*Radar Plots*

Attacking	                Defending
goals	                losses in low third
xG	                opponents shots on target
shots on target	    recoveries in high third
accurate passes	        opponents xg
possession            percentage	duels won
shots	                opponents goals

```{r}
# Data for radar plot (average values for each team)
avg_data_team = dt %>%
  group_by(team) %>%
  summarise(goals = mean(goals, na.rm = TRUE),
            xg = mean(xg, na.rm = TRUE),
            shots_on_target = mean(shots_on_target, na.rm = TRUE),
            passes_accurate = mean(passes_accurate, na.rm = TRUE),
            poss_pct = mean(poss_pct, na.rm = TRUE),
            shots = mean(shots, na.rm = TRUE),
            losses_low = mean(losses_low, na.rm = TRUE),
            rec_high = mean(rec_high, na.rm = TRUE),
            duels_won = mean(duels_won, na.rm = TRUE),
            opp.goals = mean(opp.goals, na.rm = TRUE),
            opp.xg = mean(opp.xg, na.rm = TRUE),
            opp.shots_on_target = mean(opp.shots_on_target, na.rm = TRUE))

# Compute GLOBAL min/max values (ensuring consistency across plots)
global_min_vals <- avg_data_team %>%
  summarise(across(where(is.numeric), min, na.rm = TRUE)) %>%
  mutate(team = "Min")

global_max_vals <- avg_data_team %>%
  summarise(across(where(is.numeric), max, na.rm = TRUE)) %>%
  mutate(team = "Max")

# Variables to flip (low values are better)
flip_vars <- c("losses_low", "opp.goals", "opp.xg", "opp.shots_on_target")

# Normalize function
normalize <- function(x, min_x, max_x) {
  (x - min_x) / (max_x - min_x)
}

# Normalize all values using GLOBAL min/max
normalize_data <- function(df, min_vals, max_vals, flip_vars) {
  df_norm <- df
  for (col in names(df)[-1]) {  # Exclude "team" column
    df_norm[[col]] <- normalize(df[[col]], min_vals[[col]], max_vals[[col]])

    # Flip selected defensive variables
    if (col %in% flip_vars) {
      df_norm[[col]] <- 1 - df_norm[[col]]
    }
  }
  return(df_norm)
}

# Normalize data
avg_data_team_norm <- normalize_data(avg_data_team, global_min_vals, global_max_vals, flip_vars)

# Ensure min/max values are properly scaled
global_min_vals[-ncol(global_min_vals)] <- 0  # Exclude "team" column
global_max_vals[-ncol(global_min_vals)] <- 1

# Bind datasets together
radar_data_team <- rbind(global_min_vals, global_max_vals, avg_data_team_norm)


```


```{r}
plot_radar_team <- function(radar_data, team1="Yale Bulldogs", team2){
  radar_subset = radar_data %>%
    filter(team %in% c("Min", "Max", team1, team2))
  
  # Rename variables for better readability
  var_labels = c("goals" = "Goals", "xg" = "Expected Goals (xG)", "shots_on_target" = "Shots on Target", "passes_accurate" = "Accurate Passes", "poss_pct" = "Possession Percentage", "shots" = "Shots", "losses_low" = "Losses in Low Third", "rec_high" = "Recoveries in High Third", "duels_won" = "Duels Won", "opp.goals" = "Opponent Goals", "opp.xg" = "Opponent xG", "opp.shots_on_target" = "Opponent Shots on Target")
  
  # Function to generate radar plots for a given category of variables
  create_radar <- function(vars, title) {
    radar_values <- radar_subset %>%
      select(team, all_of(vars)) %>%
      rename_with(~ var_labels[.x], .cols = all_of(vars)) %>%
      column_to_rownames(var = "team")
    
    # Convert to matrix format required by radarchart
    radar_matrix <- as.data.frame(radar_values)
    radarchart(radar_matrix, axistype = 0,
               pty = 32,
               pcol = c("red", "blue"), pfcol = c(rgb(1,0,0,0.4), rgb(0,0,1,0.4)),
               plty = c(1, 1),
               plwd = 2, title = title,
               cglcol = "gray", cglty = 1, cglwd = 0.8,  # Customize grid lines
               vlcex = 1.2 # Adjust label size
    )
  }
  
  # Create radar plots for attacking and defending variables
  par(mfrow = c(1, 2))
  create_radar(c("goals", "xg", "shots_on_target", "passes_accurate", "poss_pct", "shots"), "Attacking")
  create_radar(c("losses_low", "rec_high", "duels_won", "opp.goals", "opp.xg", "opp.shots_on_target"), "Defending")
  
  # Add legend
  legend("topright", legend = c(team1, team2), fill = c("red", "blue"), pch = 15, pt.cex = 2, bty = "n")
}
```

```{r}
plot_radar_team(radar_data_team, "Yale Bulldogs", "Princeton Tigers")
```


Radar charts using ggplot

```{r}
plot_radar_team2 <- function(dt, team1 = "Yale Bulldogs", team2) {
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(scales)
  library(ggforce)  # for coord_radar

  # --- 1. Aggregate team stats ---
  avg_data_team <- dt %>%
    group_by(team) %>%
    summarise(across(c(goals, xg, shots_on_target, passes_accurate,
                       poss_pct, shots, losses_low, rec_high,
                       duels_won, opp.goals, opp.xg, opp.shots_on_target),
                     ~ mean(.x, na.rm = TRUE)))

  # --- 2. Global min/max for normalization ---
  global_min_vals <- avg_data_team %>%
    summarise(across(where(is.numeric), min)) %>%
    mutate(team = "Min")

  global_max_vals <- avg_data_team %>%
    summarise(across(where(is.numeric), max)) %>%
    mutate(team = "Max")

  # --- 3. Normalize & flip values ---
  flip_vars <- c("losses_low", "opp.goals", "opp.xg", "opp.shots_on_target")

  normalize_data <- function(df, min_vals, max_vals, flip_vars) {
    norm_df <- df
    for (col in names(df)[-1]) {
      norm_df[[col]] <- (df[[col]] - min_vals[[col]]) / (max_vals[[col]] - min_vals[[col]])
      if (col %in% flip_vars) {
        norm_df[[col]] <- 1 - norm_df[[col]]
      }
    }
    return(norm_df)
  }

  avg_data_team_norm <- normalize_data(avg_data_team, global_min_vals, global_max_vals, flip_vars)

  # --- 4. Filter for selected teams ---
  radar_subset <- avg_data_team_norm %>%
    filter(team %in% c(team1, team2))

  # --- 5. Reshape to long format ---
  radar_long <- radar_subset %>%
    pivot_longer(-team, names_to = "metric", values_to = "value")

  # Better labels
  var_labels <- c("goals" = "Goals", "xg" = "Expected Goals (xG)", "shots_on_target" = "Shots on Target", 
                  "passes_accurate" = "Accurate Passes", "poss_pct" = "Possession %", 
                  "shots" = "Shots", "losses_low" = "Losses in Low Third", "rec_high" = "Recoveries High", 
                  "duels_won" = "Duels Won", "opp.goals" = "Opponent Goals", "opp.xg" = "Opponent xG", 
                  "opp.shots_on_target" = "Opponent Shots on Target")

  radar_long$metric_label <- factor(var_labels[radar_long$metric], levels = var_labels)

  # --- 6. Split into attacking and defending plots ---
  attacking_vars <- c("Goals", "Expected Goals (xG)", "Shots on Target", 
                      "Accurate Passes", "Possession %", "Shots")
  defending_vars <- c("Losses in Low Third", "Recoveries High", "Duels Won", 
                      "Opponent Goals", "Opponent xG", "Opponent Shots on Target")

  # --- 7. Plotting function ---
  plot_radar <- function(df, category, title) {
    ggplot(df %>% filter(metric_label %in% category), 
           aes(x = metric_label, y = value, group = team, color = team, fill = team)) +
      geom_polygon(alpha = 0.3, linewidth = 1.2) +
      geom_point(size = 3) +
      coord_polar() +
      theme_minimal(base_size = 12) +
      theme(axis.text.x = element_text(size = 10, face = "bold", angle = 5),
            panel.grid = element_line(color = "gray85"),
            legend.position = "top") +
      labs(title = title, x = NULL, y = NULL)
  }

  # --- 8. Plot side by side ---
  p1 <- plot_radar(radar_long, attacking_vars, "Attacking Metrics")
  p2 <- plot_radar(radar_long, defending_vars, "Defending Metrics")

  # Display plots side-by-side
  gridExtra::grid.arrange(p1, p2, ncol = 2)
}

```


```{r}
plot_radar_team2(dt, "Yale Bulldogs", "Princeton Tigers")
```

Another style

```{r}
plot_radar_team3 <- function(dt, team1 = "Yale Bulldogs", team2) {
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(scales)
  library(ggforce)  # for coord_radar
  library(ggshakeR)  # for plot_pizza

  data1 <- dt %>%
    filter(team == team1) %>%
    select(goals, xg, shots_on_target, passes_accurate,
           poss_pct, shots, losses_low, rec_high,
           duels_won, opp.goals, opp.xg, opp.shots_on_target)
  
  data2 <- dt %>%
    filter(team == team2) %>%
    select(goals, xg, shots_on_target, passes_accurate,
           poss_pct, shots, losses_low, rec_high,
           duels_won, opp.goals, opp.xg, opp.shots_on_target)
  # Combine data for both teams
  data <- rbind(data1, data2)
  data$team <- c(team1, team2)
  
  # Comparison plot using plot_pizza
  comp_pizza <- plot_pizza(
    data = data,
    type = "comparison", 
    template = "custom",
    player_1 = team1, 
    player_2 = team2, 
    season_player_1 = "Last 365 Days Men's Big 5 Leagues, UCL, UEL", 
    season_player_2 = "Last 365 Days Men's Big 5 Leagues, UCL, UEL",
    color_compare = "#41ab5d", 
    theme = "black"
  )

  # --- 7. Return or display the comparison plot ---
  comp_pizza
}

```


```{r}
plot_radar_team3(dt, "Yale Bulldogs", "Princeton Tigers")
```


```{r}
plot_radar_team_og <- function(dt, team1, team2) {
  
  # Data for radar plot (average values for each team)
  avg_data_team <- dt %>%
    group_by(team) %>%
    summarise(goals = mean(goals, na.rm = TRUE),
              xg = mean(xg, na.rm = TRUE),
              shots_on_target = mean(shots_on_target, na.rm = TRUE),
              passes_accurate = mean(passes_accurate, na.rm = TRUE),
              poss_pct = mean(poss_pct, na.rm = TRUE),
              shots = mean(shots, na.rm = TRUE),
              losses_low = mean(losses_low, na.rm = TRUE),
              rec_high = mean(rec_high, na.rm = TRUE),
              duels_won = mean(duels_won, na.rm = TRUE),
              opp.goals = mean(opp.goals, na.rm = TRUE),
              opp.xg = mean(opp.xg, na.rm = TRUE),
              opp.shots_on_target = mean(opp.shots_on_target, na.rm = TRUE))
  
  # Compute GLOBAL min/max values (ensuring consistency across plots)
  global_min_vals <- avg_data_team %>%
    summarise(across(where(is.numeric), min, na.rm = TRUE)) %>%
    mutate(team = "Min")
  
  global_max_vals <- avg_data_team %>%
    summarise(across(where(is.numeric), max, na.rm = TRUE)) %>%
    mutate(team = "Max")
  
  # Variables to flip (low values are better)
  flip_vars <- c("losses_low", "opp.goals", "opp.xg", "opp.shots_on_target")
  
  # Normalize function
  normalize <- function(x, min_x, max_x) {
    (x - min_x) / (max_x - min_x)
  }
  
  # Normalize all values using GLOBAL min/max
  normalize_data <- function(df, min_vals, max_vals, flip_vars) {
    df_norm <- df
    for (col in names(df)[-1]) {  # Exclude "team" column
      df_norm[[col]] <- normalize(df[[col]], min_vals[[col]], max_vals[[col]])
      
      # Flip selected defensive variables
      if (col %in% flip_vars) {
        df_norm[[col]] <- 1 - df_norm[[col]]
      }
    }
    return(df_norm)
  }
  
  # Normalize data
  avg_data_team_norm <- normalize_data(avg_data_team, global_min_vals, global_max_vals, flip_vars)
  
  min_row <- as.data.frame(t(rep(0, ncol(avg_data_team_norm) - 1)))
  names(min_row) <- names(avg_data_team_norm)[-1]
  min_row$team <- "Min"

  max_row <- as.data.frame(t(rep(1, ncol(avg_data_team_norm) - 1)))
  names(max_row) <- names(avg_data_team_norm)[-1]
  max_row$team <- "Max"

  # Bind normalized data with proper 0/1 min/max rows
  radar_data_team <- rbind(min_row, max_row, avg_data_team_norm)

  radar_subset <- radar_data_team %>%
    filter(team %in% c("Min", "Max", team1, team2))
  
  print(radar_subset)  # Check the radar_subset data frame for debugging
  
  # Rename variables for better readability
  var_labels <- c("goals" = "Goals", "xg" = "Expected Goals (xG)", "shots_on_target" = "Shots on Target", 
                  "passes_accurate" = "Accurate Passes", "poss_pct" = "Possession Percentage", 
                  "shots" = "Shots", "losses_low" = "Losses in Low Third", "rec_high" = "Recoveries in High Third", 
                  "duels_won" = "Duels Won", "opp.goals" = "Opponent Goals", "opp.xg" = "Opponent xG", 
                  "opp.shots_on_target" = "Opponent Shots on Target")
  
  # Function to generate radar plots for a given category of variables
  create_radar <- function(vars, title) {
    radar_values <- radar_subset %>%
      select(team, all_of(vars)) %>%
      rename_with(~ var_labels[.x], .cols = all_of(vars)) %>%
      column_to_rownames(var = "team")
    
    # Make sure the row order is Min, Max, team1, team2
    row_order <- c("Min", "Max", team1, team2)
    existing_rows <- intersect(row_order, rownames(radar_values))
    radar_matrix <- radar_values[existing_rows, , drop = FALSE]
    
    # Convert to matrix format required by radarchart
    radar_matrix <- as.data.frame(radar_values)
    radarchart(radar_matrix, axistype = 0,
               pty = 32,
               pcol = c("red", "blue"), pfcol = c(rgb(1,0,0,0.4), rgb(0,0,1,0.4)),
               plty = c(1, 1),
               plwd = 2, title = title,
               cglcol = "gray", cglty = 1, cglwd = 0.8,  # Customize grid lines
               vlcex = 1 # Adjust label size
    )
  }
  
  # Create radar plots for attacking and defending variables
  par(mfrow = c(1, 2), mar = c(1, 2, 2, 2), oma = c(0, 1, 2, 1))
  create_radar(c("goals", "xg", "shots_on_target", "passes_accurate", "poss_pct", "shots"), "Attacking")
  create_radar(c("losses_low", "rec_high", "duels_won", "opp.goals", "opp.xg", "opp.shots_on_target"), "Defending")
  
  # Add legend
  legend("topright", legend = c(team1, team2), fill = c("red", "blue"), pch = 15, pt.cex = 2, bty = "n")
}
```

```{r}
plot_radar_team_og(dt, "Yale Bulldogs", "Princeton Tigers")
```

```{r}
plot_radar_team_og(dt, "Harvard Crimson", "Cornell Big Red")
```


```{r}
plot_radar_team_og(dt, "Harvard Crimson", "Yale Bulldogs")
```
